Understanding Factors Contributing to DineSafe Violation Severity

### 2024-11-28 paper/paper.qmd
update qmd with new appendices

---
title: "Understanding Factors Contributing to DineSafe Violation Severity"
subtitle: "Utilizing multinomial logistic regression to visualize and understand components from June 2022 to June 2024 DineSafe data"
author: 
  - Michelle Ji
thanks: "Code and data are available at: https://github.com/michj3/dinesafe-toronto"
date: today
date-format: long
abstract: "This study aims to understand the relationship between various factors, such as minimum inspections per year and type of food establishment, and the severity of DineSafe violations in Toronto. Inspection data from June 2022 to June 2024 were compiled and analyzed through a multinomial logistic regression model. Major findings include specific types of food establishments, such as formal dining and fast food, and increasing the number of minimum inspections had higher likelihoods of a more severe infraction when compared to a baseline “minor” violation. Implications from this study can assist to raise awareness among both consumers and business owners, potentially leading to stricter adherence to health guidelines or policy changes aimed at improving and maintaining public health."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Various public health measures have recently been put into place as a means to ensure public safety and sanitation. Specific to Toronto, DineSafe is Toronto Public Health’s food inspection program and is based on provincial and municipal regulations. Each establishment that processes, manufactures, prepares, stores, or serves food must be registered for inspection. As such, the inspection process and frequency will likely depend on the type of food establishment and the level of risk associated with the food handled. Each inspection follows a set of guidelines, often assessing the food temperature control, contamination, pest control, employee hygiene, maintenance of sanitation in public areas, and waste disposal [@foodsafety]. Infraction levels are categorized by “minor”, “crucial”, and “significant”, in increasing severity order. Depending on the infraction level, results can vary from a verbal notice to closure of the establishment [@eohu]. This paper examines the factors influencing the severity of DineSafe violations, addressing a gap in public knowledge to provide useful insights for food establishments and policymakers. By understanding these factors, targeted actions can be taken to enhance public health outcomes for the citizens of Toronto.

The estimand in the paper is the relationship between factors such as the minimum number of inspections per year and the type of food establishment on the severity of a DineSafe violation. By analyzing the dataset and developing a multinomial logistic regression model, the primary objective is to identify the factors that influence the severity levels of DineSafe violations. This knowledge can raise awareness and guide employers, employees and policymakers to adopt improved habits or implement changes in policies. One main finding is that specific food establishments, such as formal dining, fast food, and industrial food preparation places, had a higher likelihood of having a “crucial” violation over a “minor” violation. This can lead to serious consequences, such as an increased likelihood of establishment closure. Therefore, it is important for business owners to be aware of these factors so they can take proactive measures to address potential issues and violations.

This paper is organized further into five main sections. @sec-data discusses how the dataset was measured and important variables of interest. @sec-model describes the model used for analysis and potential weaknesses. @sec-results explores the main findings of the model and @sec-discussion highlights real world implications and possible future courses of action. @sec-appendices discusses more in-depth how the data was cleaned for modeling and a discussion on observational data in this dataset, emphasizing its significance and the implications of the findings. @sec-appendices also includes a datasheet for the dataset.

@@

The dataset was obtained through the City of Toronto’s OpenData Library [@opendatatoronto] and is titled “DineSafe” [@data]. As part of Toronto Public Health’s food safety program, DineSafe is a program that inspects all establishments preparing and serving food; these types of establishments range from food trucks to sit-down restaurants. Municipal health departments conduct inspections that are either routine, complaint-driven, or follow-up checks. Establishments will be evaluated on predefined criteria, such as cleanliness, food handling, and storage.  Each inspection will result in a pass, conditional pass, or closed notice depending on the severity level of any violations and courses of action. Results, including the number and type of violations, are logged into Toronto’s municipal database, reflected in the DineSafe dataset through OpenDataToronto.

The dataset includes data containing inspection details of food establishments from June 2022 to June 2024. There were no other similar datasets. @tbl-1 details the first few observations of the cleaned dataset for analysis.

@@

The severity level variable is a categorical variable with three levels: “C” for crucial, “S” for significant, and “M” for minor. The most severe infraction is “crucial” and the least severe is “minor”. Each level corresponds to the level of sanitation violation that the food establishment made and will correspond with different courses of action and outcome. @fig-1 details the distribution of the different infraction levels. From @fig-1, the most frequent type of infraction level is "minor" and "crucial" infractions of regulations are the least frequent. From a public health perspective, this is a positive sign, as it suggests that most food establishments maintain relatively high standards of sanitation. This is reassuring for consumers, as it indicates that food establishments are generally safe and comply with health regulations.

@@

ggplot(top_severity, aes(x = reorder(severity, -n), y = n)) + 
  geom_bar(stat = "identity", fill = "springgreen4", color = "black") + 
  labs(
    x = "Frequency of Dinesafe Violation Severity Level", 
    y = "Frequency"
  ) + 
  theme_minimal()
```
The variable focusing on the minimum number of inspections per year is a quantitative variable that ranges from 1 to 3 visits per year. Typically, the number of inspections will vary depending on the specific type of establishment, the food preparation processes, volume and type of food served and other related criteria [@aboutdinesafe]. @fig-2 details the distribution of the number of inspections per year the food establishments in the dataset receive. @fig-2 highlights that most commonly, food establishments in Toronto receive around 3 inspections per year and rarely will receive only 1. This may relate to the complexity of certain restaurants or food establishments, which often require more thorough inspections due to multiple components, such as kitchen operations, food preparation, and serving processes. Generally, and rarely, only very basic establishments receive a single inspection per year.

@@

@tbl-2 details the mean and median number of inspections of Toronto food establishment inspections. The variable focusing on food establishment type is a qualitative variable that contains the types of food establishments inspected, which include food trucks, serving kitchens, bar and cocktail lounges, and restaurants, to name a few.

@@

```{r}
#| label: fig-3
#| fig-cap: "Most to least frequented food establishment types inspected through DineSafe in Toronto"
#| echo: false

@@

A multinomial logistic regression model was used to determine if the severity of a DineSafe violation had a relationship with the type of food establishment and minimum number of inspections per year an establishment receives. A multinomial logistic regression model is used on nominal outcome variables, in which the log odds of the outcomes are modeled as a linear combination of the predictor variables [@aboutmodel]. Since the severity variable has multiple levels and is not dichotomous, a multinomial logistic regression model is used instead of a binary logistic model. Coefficients correspond to the log-odds of being in a specific category (either "C - Crucial" or "S - Significant") relative to the reference category ("M - Minor"). Log odds is a way to describe the likelihood of an event happening; in this case, the event would be a different severity level relative to the “minor” severity level that uses a logarithmic scale.

@@

```{=latex}
\begin{align}
\log\left(\frac{P(\text{Severity}= \text{Significant})}{P(\text{Severity} = \text{Minor})} \right) &= 
\beta_{0,1} + \beta_{1,1} \times \text{min inspections} + \\
& \beta_{2,1} \times \text{establishmenttype:Fast Food} + \\
& \beta_{3,1} \times \text{establishmenttype:Formal Dining} + \\
& \beta_{4,1} \times \text{establishmenttype:Grocery} + \\
& \beta_{5,1} \times \text{establishmenttype:Industrial} + \\
& \beta_{6,1} \times \text{establishmenttype:Institution Food Prep}
\end{align}

@@

To check the validity of the model used, it’s important to check for independence of observations, linearity in log-odds, multicollinearity, and sufficient sample size. For independence of observations, independent observations are assumed since each observation in the dataset is a separate restaurant or individual violation of DineSafe. They are treated as separate violations or inspections and are unaffected by other inspections. One inspection’s outcome does not influence another establishment’s outcome and each inspection for the same establishment is treated as a stand-alone event or observation. To look at linearity in log-odds, we can use the numerical predictor variable, the minimum number of inspections per year for an establishment, and check the linearity of the relationship between that and the log-odds of the predicted probabilities from the multinomial model. This is shown in @fig-4, where there is a general linear relationship between the two variables.

@@

plot(dinesafe_data$min_inspections, log_odds, 
     xlab = "min_inspections", ylab = "Predicted Log-Odds")
```
One thing to note is the distinct observations that form almost vertical lines. This should not be of concern, as there are only three possible values of minimum inspections and are integers: 1, 2, and 3. The overall trend of this plot is linear, which we can conclude this assumption holds. To check for multicollinearity, we need to first convert the food establishment type variable into a dummy variable, holding it as a “numerical” value instead of categorical. Then, we can use the cor() function to look at correlation and thus, multicollinearity. The values of correlation between the variables are in @fig-8 in @sec-appendices-figures. The variables do not seem to have high correlation with each other, indicating that our model likely does not have severe multicollinearity and its conclusions can be treated as valid. Finally, it’s important to note that for a model to be valid, there needs to be sufficient sample size. Our model uses a dataset [@data] that has over 18,000 observations, so sample size is not a concern for model validity.

@@

Model results are summarized in @tbl-3. The labels of "x 1" and "x 2" next to each of the predictor variables indicates what level severity we are evaluating the model at compared to the reference level, "minor". "x 1" indicates comparing "crucial" to "minor", whereas "x 2" indicates comparing "signficant" to "minor".

@@

modelsummary(dinesafe_model, stars = FALSE, gof_omit = ".*")
```

@fig-6 illustrates the coefficients and their corresponding 95% confidence intervals for the predictor variables in the multinomial logistic regression model as a forest plot. A 95% confidence interval (CI) is a statistical range that is likely to contain the true value of a population parameter, in this case the regression coefficient, with 95% certainty. The forest plot allows visualization of the odds ratios with the confidence intervals, making it easier to interpret which predictors have a significant impact across different outcome classes. If the confidence interval for a coefficient includes 0, or the p-value is greater than 0.05, it suggests that there is no statistically significant effect of that predictor (establishment type or minimum inspections) on the response variable (severity) at the chosen level of significance of 0.05.

@@

@@

From the model analysis, it is evident that a greater number of minimum inspections per year is associated with a higher likelihood of a significant violation. This finding carries several important implications for both public health and business operations. First, it can encourage business owners to minimize violations, as each follow-up inspection counts towards the total number of inspections. An increase in inspections or violations may result in the closure of an establishment, which not only negatively affects the business but also represents a loss of investment for the stakeholders involved. Furthermore, since businesses are required to display their DineSafe inspection results publicly, frequent negative changes in the rating can deter customers, leading to reduced patronage and further financial loss. Additionally, this insight may prompt businesses to implement more stringent sanitation practices and provide enhanced public health training for employees. Although these measures may be short-term costs, they can prove to be beneficial in the long run if they help prevent violations that can result in higher costs, thereby maintaining a good standing in public health inspections. From a public health perspective, the findings highlight the importance of motivating employees to adhere to health guidelines and sanitation protocols, as this not only helps reduce the likelihood of severe violations but also fosters a healthier work environment. This approach can contribute to more consistent compliance with health regulations, benefiting both employees and the business as a whole.

@@

The model suggests that formal dining, fast food, and institutional food preparation establishments are more likely to receive "crucial" ratings when compared to "minor" violations. For establishment owners and employees, this finding is particularly significant, as it highlights the importance of maintaining high standards of sanitation in these types of establishments. These establishments typically handle more raw food, such as meats and fresh produce, which increases the potential for foodborne contamination. Therefore, strict adherence to public health guidelines and proper food preparation practices is important in ensuring the safety of both employees and customers.

Improper handling of raw foods can lead to the spread of foodborne illnesses, such as salmonella and E. coli, which pose significant public health risks. A disease outbreak traced back to an establishment can severely damage its reputation, leading to public distrust and a decline in customers. The public health consequences are severe, but so are the business repercussions, potentially resulting in a loss of customers and revenue. Given these risks, employees in these establishments should be especially strict about sanitation practices and adher to established health protocols. Awareness of the potential severity of violations may encourage staff to prioritize hygiene and food safety. Additionally, this information can drive the development of more vigorous employee training programs focused on sanitation practices and food safety. Not only will these programs help prevent severe violations and improve public health outcomes, but they will also positively impact business operations. Maintaining a strong DineSafe rating will be visible to the public, and establishments with higher ratings are likely to attract more customers, improving their overall business success. Therefore, investing in better sanitation practices is beneficial for both public health and long-term business sustainability.

The findings not only have implications for business owners but also for the inspectors responsible for monitoring food safety. Given the identified association between certain types of establishments—such as formal dining, fast food, and institutional food preparation—and a higher likelihood of receiving severe DineSafe violations, inspectors can use this information to adopt a more strict approach when inspecting these establishments. By recognizing that these types of businesses are more prone to severe violations, inspectors can focus their attention more closely on areas of higher risk, such as the handling of raw meats and produce, food storage, and sanitation practices. More rigorous inspection process not only minimizes the chances of a serious health violation occurring but also strengthens the overall public health framework by ensuring that establishments are held accountable for maintaining high standards of cleanliness and food safety. This can reinforce trust in the regulatory system and ensure that the DineSafe program remains a reliable indicator of a restaurant's commitment to health and safety.

@@

While the results provide important insights for addressing public health practices in food establishments, the DineSafe model and its implications are specific to Toronto, as the data used in the analysis was collected from this city. Caution should be exercised when attempting to generalize these findings to other cities, as different regions may yield varying outcomes. Additionally, it is important to note that the model identifies associations between variables but does not establish causation, and should not be used as cause-and-effect ties. Moreover, potential biases in the data collection process cannot be ruled out. For example, individual inspector tendencies—such as one inspector being consistently stricter than others—could influence the severity of violations assigned, potentially skewing the results.

While all cities have similar programs to inspect public health and sanitation measures, one potential future step is to create more universal policies across a given province or country. More in-depth analysis could be done when practices and guidelines are standarized, providing more causation links that can be used to improve a region’s public health. Also, public health initiatives and educational campaigns could be developed specifically for establishments that are prone to severe violations, educating both owners and employees on best practices for food safety. This could include targeted training on handling raw foods, sanitation, and compliance with food safety regulations. On the consumer side, making DineSafe data more accessible to the public, such as through a website or app, may also be a future step worth considering as a way to encourage consumers to make more informed decisions and incentivize businesses to strive for better performance.

@@

## Observational Data in the Context of DineSafe {#sec-appendices-observational}

@@

Observational datasets generally capture only correlations rather than causal relationships, making it difficult to definitively state that one factor directly influences the severity of violations. In the context of DineSafe, it is difficult to state that minimum inspections per year or type of establishment has a true casual relationship with the severity of a violation, although there is a statisticlaly significant correlation. These limitations require careful interpretation of results and underscore the importance of supplementary methods, such as experimental designs or in-depth qualitative research, to validate findings and address areas that are not as well understood. However, that also poses an issue for DineSafe. Ethically, it would not be feasible to conduct experimental tests, such as A/B testing. the potential manipulation of inspection rigor between groups could lead to inconsistencies in public health protection. For example, if one group of establishments receives stricter inspections than another, consumers dining at less rigorously inspected establishments may face greater health risks that cannot be taken accountability for. Often, A/B testing requires double-blind experiments to account for any biases, raising an ethical concern for public food safety; businesses and consumers might unknowingly participate in an experimental setup, violating principles of informed consent, damaging trust in public health authorities and the integrity of the inspection system.

## Idealized Survey {#sec-appendices-idealized}

Another way to assess DineSafe as a public health program is through a consumer’s perspective. In 2020, a study was conducted to evaluate the effects of DineSafe’s disclosure system to customers [@survey]. In this study, a survey was conducted and advertised via social media platforms. The study consisted of 25 respondents of different age classes, all above the age of 18. While this survey gathered important data to analyze consumer’s knowledge on DineSafe, the sample size is fairly small to draw any definitive conclusions. Thus, in an idealized world with a high budget, a similar survey could be conducted at a larger and more representative scale. The idealized methodology for this survey to analyze consumer knowledge about DineSafe could involve quota sampling, with probability proportionate to the geographical location according to population data. For example, we could sample the Greater Toronto Area, however, this method ensures that the number of people polled is proportional to amount of people living in that region who would eat at those food establishments. Those in Downtown Toronto would have a higher probability of accessing and answering the survey compared to those in a less populated, uptown area.

After the sample has been selected with the above methods, the poll itself will be conducted through a survey sent to the chosen participants to reduce interviewer bias, or the effect of the interviewer’s views on the measured responses, which can affect results when face-to-face or phone interviews are utilized [@citeTB]. To also reduce the social desirability effect, the survey must emphasize in the beginning that the respondents’ identity will be kept anonymous to the researchers and the final reported data [@citeStantcheva]. To encourage participation, a financial incentive should be offered for full completion of the poll. Displaying the progress bar and indicating the survey length at the very beginning also helps encourage participants to complete the full questionnaire. Our poll would also be submitted to the appropriate ethics review boards, likely to those involved with Toronto Public Health, to increase confidence in our methodology.

@@

>>>>

---
title: "Understanding Factors Contributing to DineSafe Violation Severity"
subtitle: "Multinomial logistic regression reveals higher severity violations in formal dining and fast food establishments from June 2022 to June 2024"
author: 
  - Michelle Ji
thanks: "Code and data are available at: https://github.com/michj3/dinesafe-toronto"
date: today
date-format: long
abstract: "This study aims to understand the relationship between various factors, such as minimum inspections per year and type of food establishment, and the severity of DineSafe violations in Toronto. Inspection data from June 2022 to June 2024 were compiled and analyzed through a multinomial logistic regression model. Major findings include specific types of food establishments, such as formal dining and fast food, and increasing the number of minimum inspections had higher likelihood of more severe infraction. Implications from this study can assist to raise awareness among both consumers and business owners, potentially leading to stricter adherence to health guidelines or policy changes aimed at improving and maintaining public health."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Various public health measures have recently been put into place to ensure public safety and sanitation. DineSafe is Toronto Public Health’s food inspection program and is based on provincial and municipal regulations. Each establishment that processes, manufactures, prepares, stores, or serves food must be registered for inspection. As such, the inspection process and frequency will likely depend on the level of risk associated with the food handled. Each inspection follows a set of guidelines, often assessing the food temperature control, contamination, pest control, employee hygiene, maintenance of sanitation in public areas, and waste disposal [@foodsafety]. Depending on the infraction level, results can vary from a verbal notice to closure of the establishment [@eohu]. This paper examines factors influencing the severity of DineSafe violations, providing useful findings for food establishments and policymakers. By understanding these factors, targeted actions can be taken to enhance public health outcomes for the citizens of Toronto.

The estimand in the paper is the relationship between factors such as the minimum number of inspections per year and the type of food establishment on the severity of a DineSafe violation. By analyzing the dataset and developing a multinomial logistic regression model, the primary objective is to identify the factors that influence the severity levels of DineSafe violations. This knowledge can raise awareness and guide employers, employees and policymakers to adopt improved habits or implement changes in policies. One main finding is that specific food establishments, such as formal dining, fast food, and industrial food preparation places, had a higher likelihood of having a “crucial” violation over a “minor” violation. This can lead to serious consequences, such as an increased likelihood of establishment closure. Therefore, it is important for business owners to be aware of these factors so they can take proactive measures.

This paper is organized further into five main sections. @sec-data discusses how the dataset was measured and important variables of interest. @sec-model describes the model used for analysis and potential weaknesses. @sec-results explores the main findings of the model and @sec-discussion highlights real world implications and possible future courses of action. @sec-appendices discusses more in-depth data cleaning, a discussion on observational data in this dataset, analysis of a previously conducted survey and an idealized survey, and connections to pre-existing literature. @sec-appendices also includes a datasheet for the dataset.

@@

The dataset was obtained through the City of Toronto’s OpenData Library [@opendatatoronto] and is titled “DineSafe” [@data]. As part of Toronto Public Health’s food safety program, DineSafe is a program that inspects all establishments preparing and serving food; these types of establishments range from food trucks to sit-down restaurants. Municipal health departments conduct inspections that are either routine, complaint-driven, or follow-up checks. Establishments will be evaluated on predefined criteria, such as cleanliness, food handling, and storage. Each inspection will result in a pass, conditional pass, or closed notice depending on the severity level of any violations and courses of action. Routine inspections typically occur once to three times a year. Follow up inspections will typically occur within 2 weeks of the initial inspection. Results, including the number and type of violations, are logged into Toronto’s municipal database, reflected in the DineSafe dataset through OpenDataToronto.

The dataset includes data containing inspection details of food establishments from June 2022 to June 2024. There were no other similar datasets to consider. @tbl-1 details the first few observations of the cleaned dataset for analysis.

@@

### Response Variable {#sec-data-variables-response}

The severity level variable is a categorical variable with three levels: “C” for crucial, “S” for significant, and “M” for minor. The most severe infraction is “significant” and the least severe is “minor”. Each level corresponds to the level of sanitation violation that the food establishment made and will correspond with different courses of action and outcome. @fig-1 details the distribution of the different infraction levels. From @fig-1, the most frequent type of infraction level is "minor" and "crucial" infractions of regulations are the least frequent. From a public health perspective, this is a positive sign, as it suggests that most food establishments maintain relatively high standards of sanitation. This is reassuring for consumers, indicating food establishments generally comply with health regulations.

@@

ggplot(top_severity, aes(x = reorder(severity, -n), y = n)) + 
  geom_bar(stat = "identity", fill = "springgreen4", color = "black") + 
  labs(
    x = "Frequency of Dinesafe Violation Severity Level", 
    y = "Frequency"
  ) + 
  theme_minimal()
```
\newpage

### Predictor Variables {#sec-data-variables-predictors}

There are two predictor variables in the model, minimum inspections and food establishment type. The variable focusing on the minimum number of inspections per year is a quantitative variable that ranges from 1 to 3 visits per year. Typically, the number of inspections will vary depending on the specific type of establishment, the food preparation processes, volume and type of food served and other related criteria [@aboutdinesafe]. @fig-2 details the distribution of the number of inspections per year the food establishments in the dataset receive. @fig-2 highlights that most commonly, food establishments in Toronto receive around 3 inspections per year and rarely receive only 1. This may relate to the complexity of certain restaurants or food establishments, which often require more thorough inspections due to multiple components, such as kitchen operations, food preparation, and serving processes.

@@

@tbl-2 details the mean and median number of inspections of Toronto food establishment inspections.

@@

\newpage
```{r}
#| label: fig-3
#| fig-cap: "Most to least frequented food establishment types inspected through DineSafe in Toronto"
#| echo: false

@@

Using the statistical programming software R [@citeR] and VGAM package [@vgam], a multinomial logistic regression model was used to determine if the severity of a DineSafe violation had a relationship with the type of food establishment and minimum number of inspections per year an establishment receives. A multinomial logistic regression model is used on nominal outcome variables, in which the log odds of the outcomes are modeled as a linear combination of the predictor variables [@aboutmodel]. Since the severity variable has multiple levels and is not dichotomous, a multinomial logistic regression model is used instead of a binary logistic model. Coefficients correspond to the log-odds of being in a specific category (either "C - Crucial" or "S - Significant") relative to the reference category ("M - Minor"). Log odds is a way to describe the likelihood of an event happening; in this case, the event would be a different severity level relative to the “minor” severity level that uses a logarithmic scale.

@@

```{=latex}
\begin{align}
\log\left(\frac{P(\text{Severity}= \text{Crucial})}{P(\text{Severity} = \text{Minor})} \right) &= 
\beta_{0,1} + \beta_{1,1} \times \text{min inspections} + \\
& \beta_{2,1} \times \text{establishmenttype:Fast Food} + \\
& \beta_{3,1} \times \text{establishmenttype:Formal Dining} + \\
& \beta_{4,1} \times \text{establishmenttype:Grocery} + \\
& \beta_{5,1} \times \text{establishmenttype:Industrial} + \\
& \beta_{6,1} \times \text{establishmenttype:Institution Food Prep}
\end{align}

@@

To check the validity of the model used, we check independence of observations, linearity in log-odds, multicollinearity, and sufficient sample size.

For independence of observations, independent observations are assumed since each observation in the dataset is a separate restaurant or individual violation of DineSafe. They are treated as separate violations or inspections and are unaffected by other inspections. One inspection’s outcome does not influence another establishment’s outcome and each inspection for the same establishment is treated as a stand-alone event.

To look at linearity in log-odds, we can use the numerical predictor variable, the minimum number of inspections per year for an establishment, and check the linearity of the relationship between that and the log-odds of the predicted probabilities from the multinomial model. This is shown in @fig-4, where there is a general linear relationship between the two variables.

@@

plot(dinesafe_data$min_inspections, log_odds, 
     xlab = "min_inspections", ylab = "Predicted Log-Odds")
```
One thing to note is the distinct observations that form almost vertical lines. This is not of concern, as there are only three distinct values of minimum inspections and are integers: 1, 2, and 3. The overall trend of this plot is linear, which we can conclude this assumption holds.

To check for multicollinearity, the food establishment type variable was converted into a dummy variable, holding it as a “numerical” value instead of categorical. Using the cor() function to look at correlation and thus, multicollinearity, the values of correlation between the variables are in @fig-8 in @sec-appendices-figures. The variables do not seem to have high correlation with each other, indicating that our model likely does not have severe multicollinearity and its conclusions can be treated as valid.

Finally, it’s important to note that for a model to be valid, there needs to be sufficient sample size. This model uses a dataset [@data] that has over 18,000 observations, so sample size is not a concern for model validity.

@@

Model results are summarized in @tbl-3. The labels of "x 1" and "x 2" next to each of the predictor variables indicates what level severity we are evaluating the model at compared to the reference level, "minor". "x 1" indicates comparing "crucial" to "minor", whereas "x 2" indicates comparing "significant" to "minor".

@@

modelsummary(dinesafe_model, stars = FALSE, gof_omit = ".*")
```
@fig-6 illustrates the coefficients and their corresponding 95% confidence intervals for the predictor variables in the multinomial logistic regression model as a forest plot. A 95% confidence interval (CI) is a statistical range that is likely to contain the true value of a population parameter, in this case the regression coefficient, with 95% certainty. The forest plot allows visualization of the odds ratios with the confidence intervals, making it easier to interpret which predictors have a significant impact across different outcome classes. If the confidence interval for a coefficient includes 0, or the p-value is greater than 0.05, it suggests that there is no statistically significant effect of that predictor (establishment type or minimum inspections) on the response variable (severity) at the chosen level of significance of 0.05.

@@

In this paper, a multinomial logistic regression model was used to understand the relationship between the severity of a Dinesafe violation and two predictors, minimum inspections per year and type of food establishment. Notable findings are discussed in the following subsections, along with their implications.

@@

From model analysis, it is evident that a greater number of minimum inspections per year is associated with a higher likelihood of a significant violation. This finding carries several important implications for both public health and business operations.

First, it can encourage business owners to minimize violations, as each follow-up inspection counts towards the total number of inspections. An increase in inspections or violations may result in the closure of an establishment, which not only negatively affects the business but also represents a loss of investment for the stakeholders involved. Furthermore, since businesses are required to display their DineSafe inspection results publicly, frequent negative changes in the rating can deter customers, leading to further financial loss.

Additionally, this may prompt businesses to implement more stringent sanitation practices and provide better public health training for employees. Although these measures may be short-term costs, they can prove to be beneficial in the long run if they help prevent violations that can result in higher costs, thereby maintaining a good standing in public health inspections. From a public health perspective, the findings highlight the importance of motivating employees to adhere to health guidelines and sanitation protocols, as this not only helps reduce the likelihood of severe violations but also fosters a healthier work environment. This approach can contribute to more consistent compliance with health regulations, benefiting both employees and the business as a whole.

@@

The model suggests that formal dining, fast food, and institutional food preparation establishments are more likely to receive "crucial" ratings when compared to "minor" violations. For establishment owners and employees, this finding is particularly significant, highlighting the importance of maintaining high standards of sanitation in these types of establishments. These establishments typically handle more raw food, such as meats and fresh produce, which increases the potential for foodborne contamination. Therefore, strict adherence to public health guidelines and proper food preparation practices is important in ensuring the safety of both employees and customers.

Improper handling of raw foods can lead to the spread of foodborne illnesses, such as salmonella and E. coli, which pose significant public health risks. A disease outbreak traced back to an establishment can severely damage its reputation, leading to public distrust and a decline in customers. The public health consequences are severe, but so are the business repercussions, potentially resulting in a loss of customers and revenue. Given these risks, employees in these establishments should be especially strict about sanitation practices. Awareness of the potential severity of violations may encourage staff to prioritize hygiene and food safety. Additionally, this information can drive the development of more vigorous employee training programs focused on sanitation practices and food safety. Not only will these programs help prevent severe violations and improve public health outcomes, but they will also positively impact business operations. Maintaining a strong DineSafe rating will be visible to the public, and establishments with higher ratings are likely to attract more customers, improving their overall business success. Therefore, investing in better sanitation practices is beneficial for both public health and long-term business sustainability.

The model's findings not only have implications for business owners but also for the inspectors responsible for monitoring food safety. Given the identified association between certain types of establishments—such as formal dining, fast food, and institutional food preparation—and a higher likelihood of receiving severe DineSafe violations, inspectors can use this information to adopt a more strict approach when inspecting these establishments. By recognizing that these types of businesses are more prone to severe violations, inspectors can focus their attention more closely on areas of higher risk, such as the handling of raw meats and produce, food storage, and sanitation practices. More rigorous inspection process not only minimizes the chances of a serious health violation occurring but also strengthens the overall public health framework by ensuring that establishments are held accountable for maintaining high standards of cleanliness and food safety. This can reinforce trust in the regulatory system and ensure that the DineSafe program remains a reliable indicator of a restaurant's commitment to health and safety.

@@

While the results of the model provide important findings for addressing public health practices in food establishments, the DineSafe model and its implications are specific to Toronto, as the data used in the analysis was collected from this city. Caution should be exercised when attempting to generalize these findings to other cities, as different regions may yield varying outcomes. Additionally, it is important to note that the model identifies associations between variables but does not establish causation, and should not be used as cause-and-effect ties. Moreover, potential biases in the data collection process cannot be ruled out. For example, individual inspector tendencies—such as one inspector being consistently stricter than others—could influence the severity of violations assigned, potentially skewing the results.

While all cities have similar programs to inspect public health and sanitation measures, one potential future step is to create more universal policies across a given province or country. More in-depth analysis could be done when practices and guidelines are standardized, providing more causation links that can be used to improve a region’s public health. Also, public health initiatives and educational campaigns could be developed specifically for establishments that are prone to severe violations, educating both owners and employees on best practices for food safety. This could include targeted training on handling raw foods, sanitation, and compliance with food safety regulations. On the consumer side, making DineSafe data more accessible to the public, such as through a website or app, may also be a future step worth considering as a way to encourage consumers to make more informed decisions and incentivize businesses to strive for better performance.

@@

## Surveys, Sampling, and Observational data {#sec-appendices-sso}

### Observational Data in the Context of DineSafe {#sec-appendices-observational}

@@

Observational datasets generally capture only correlations rather than causal relationships, making it difficult to definitively state that one factor directly influences the severity of violations. In the context of DineSafe, it is difficult to state that minimum inspections per year or type of establishment has a true casual relationship with the severity of a violation, although there is a statistically significant correlation. These limitations require careful interpretation of results and underscore the importance of supplementary methods, such as experimental designs or in-depth qualitative research, to validate findings and address areas that are not as well understood. However, that also poses an issue for DineSafe. Ethically, it would not be feasible to conduct experimental tests, such as A/B testing. the potential manipulation of inspection rigor between groups could lead to inconsistencies in public health protection. For example, if one group of establishments receives stricter inspections than another, consumers dining at less rigorously inspected establishments may face greater health risks that cannot be taken accountability for. Often, A/B testing requires double-blind experiments to account for any biases, raising an ethical concern for public food safety; businesses and consumers might unknowingly participate in an experimental setup, violating principles of informed consent, damaging trust in public health authorities and the integrity of the inspection system.

### Analysis of Previously a Conducted Survey on DineSafe {#sec-appendices-limiations}

Another way to assess DineSafe as a public health program is through a consumer’s perspective. In 2020, a study was conducted to evaluate the effects of DineSafe’s disclosure system to customers [@survey]. In this study, a cross-sectional survey was conducted and distributed on Facebook through an anonymous third-party account. The study consisted of 25 respondents of different age classes, all above the age of 18.

One strength of this survey is its specificity to real world events. Specifically, the survey was conducted during the COVID-19 pandemic and survey questions were tailored specific to the pandemic. Researchers were able to gain insights into how public health measures and pandemic-related concerns impacted dining habits. This targeted approach enhances the relevance and applicability of their findings, providing valuable data for businesses and policymakers adapting to pandemic-related changes in consumer behavior. The consistency of the survey questions across all respondents ensures standardized responses, allowing researchers to make definitive and easy comparisons and conclusions. Specific to the pandemic, online distribution of the survey provided a method that both faciliated wider distribution of the survey and adherence to public health guidelines, protecting the physical health of respondents without exposure to COVID-19.

While this survey gathered important data to analyze consumer’s knowledge on DineSafe, there are a few limitations that can be addressed in an idealized survey. First, the sample size of 25 responses is fairly small to draw any definitive conclusions about causation or trends. It is difficult to conclude any definitive finding, as 25 responses may result in specific answers due to chance. Additionally, the process of survey distribution introduces biases that could be otherwise avoided. The survey is subject to selection bias, as Facebook users are not representative of an entire population. Specific social media platforms tend to skew to certain age groups; Facebook's largest audience is between 25 and 34 years old [@statistica], a very small subset of a population. People who do not use Facebook or have limited internet access are automatically excluded from participating, creating a biased sample. Self selection bias is also present in the conducted survey by @survey. More active users on Facebook may see the survey post or advertisement over users that do not frequently use the platform, resulting in over-representation of a subset of the population. Voluntary participation is also an issue, where people who choose to respond may have specific motivations or interests related to the survey topic, resulting in a biased response. Facebook or other social media platform's algorithm of showing posts to their users also may create additional bias, as their algorithms are designed to show content to people who are likely to share similar demographics, beliefs, or behaviors. This can result in a homogeneous sample with limited diversity.

### Idealized Survey {#sec-appendices-own}

In an idealized world with a high budget, a similar survey could be conducted at a larger and more representative scale. The idealized methodology for this survey to analyze consumer knowledge about DineSafe could involve quota sampling, with probability proportionate to the geographical location according to population data. For example, we could sample the Greater Toronto Area, however, this method ensures that the number of people polled is proportional to amount of people living in that region who would eat at those food establishments. Those in Downtown Toronto would have a higher probability of accessing and answering the survey compared to those in a less populated, uptown area.

After the sample has been selected with the above methods, the poll itself will be conducted through a survey sent to the chosen participants to reduce interviewer bias, or the effect of the interviewer’s views on the measured responses, which can affect results when face-to-face or phone interviews are utilized [@citeTB]. To reduce selection or voluntary response bias, the poll will not be distributed via social media platforms and instead, directly sent to participants. While this does not completely eliminate voluntary response bias, it will help reduce its effect on the overall response. To also reduce the social desirability effect, the survey must emphasize in the beginning that the respondents’ identity will be kept anonymous to the researchers and the final reported data [@citeStantcheva]. To encourage participation, a financial incentive should be offered for full completion of the poll. Displaying the progress bar and indicating the survey length at the very beginning also helps encourage participants to complete the full questionnaire. Our poll would also be submitted to the appropriate ethics review boards, likely to those involved with Toronto Public Health, to increase confidence in our methodology.

@@

## Linkages to Published Literature

A 2014 study investigated food inspection reports in Florida's food establishments [@florida]. Findings revealed chain restaurants had 26% fewer critical violations, benefiting from better quality control systems. Independent restaurants dominated violations, with notable issues in high-tourism districts like Miami and Jacksonville. This study's findings are highly relevant to Toronto's DineSafe program, given the city's high tourism profile. Toronto, like Miami and Jacksonville, must prioritize strict food safety practices to maintain public trust and health. The results of both @florida and the model in this paper suggest a shift in focus to restaurant-style food establishments. While independent restaurants in Florida were identified as the most frequent violators of food safety guidelines, this trend may not entirely align with Toronto’s food market dynamics. Nonetheless, it is a consideration for Toronto Public Health to keep more strict. Adapting strategies to the local context while drawing lessons from Florida’s experience can support Toronto’s efforts to maintain rigorous food safety standards in a high-tourism environment.
